---
title: "OxoScan - Figure 2"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

# Quantitative profiling of plasma glycopeptides with OxoScan


```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = T,
                      echo = F,
                      error = F,
                      warning = F,
                      cache = T,
                      dev = "png",
                      dpi = 300,
                      fig.path = paste0(getwd(),
                                        "/Plots/",
                                        gsub(Sys.Date(), ## Make folder for date of running
                                             pattern = "-", 
                                             replacement = ""),
                                        "_Plots/"))

library(data.table)
library(fs)
library(tidyverse)
library(gridExtra)
library(magrittr)
library(cowplot)
### colour palettes
library(viridisLite)
library(PNWColors)
library(viridis)
library(circlize)


## Directories
root.dir <- getwd() ## Change as necessary to enclosing folder with script (from which /Data and /Plots are accessed)

raw.data.dir <- paste0(root.dir, "/Data")

plot.output.dir <- paste0(root.dir, "/Plots/")


```

``` {R Session info}

#sessioninfo::session_info()

```

#### Read in deglycosylation experiment data (direct from DIA-NN `--extract` function)

``` {R Read in Deglycosylation DIA-NN Output, echo = T, eval = T}

## Read in treated/untreated plasma digests
deglyco.non.treated <- dir_ls(raw.data.dir, regexp = "untreated") %>% ## No glycosidase treatment
  fread() %>%
  mutate(Exp = "- Glycosidase")

deglyco.treated <- dir_ls(raw.data.dir, regexp = "16h") %>% ## + Glycosidase treatment
  fread() %>% 
  mutate(Exp = "+ Glycosidase") ## Add experiment condition

deglyco.raw <- rbind(deglyco.non.treated,
                     deglyco.treated)
  
                     
  
  

## Format longer, remove zero values, scale and add labels by each ion 
deglyco <- deglyco.raw %>%
  mutate(mz = ((Window.Low+Window.High)/2)) %>% ## Use middle of each m/z bin for precursor m/z plotting
  select(c('RT', 'mz', 'Exp', '204.087', '366.14', '274.092')) %>% 
  pivot_longer(cols = 4:6, 
               names_to = "Ion", 
               values_to = "Intensity") %>% ## Make 'long' table
  filter(Intensity != "0") %>% ## Remove 0 values
  group_by(Ion) %>% ## Scale intensity across control/deglycosidase experiments by each ion
  mutate(Scaled.Intensity = Intensity/max(Intensity)) %>%
  ungroup() %>%
  mutate(Ion.Label = ifelse(Ion == 274.092, ## Add labels describing oxonium ion (i.e. sugar monomer they derive from)
                            yes = "NeuAc",
                            no = ifelse(Ion == 204.087,
                                             yes = "HexNAc",
                                             no = "Hex+HexNAc")),
         Exp = factor(Exp, levels = c("- Glycosidase", "+ Glycosidase"))) ## As factor for plotting

#write_csv(deglyco %>%
#            mutate(Exp = as.character(Exp),
#                   Exp = ifelse(Exp == "- Glycosidase",
#                                yes= "Without Glycosidase",
#                                no = "With Glycosidase")), "SourceData/Figure2b_Figure2c.csv")
```

#### Oxonium ion map of plasma tryptic digest - Figure 2b

``` {R Figure2b_OxoniumMap, eval = T, echo = T, results = F}

## Just show oxonium ion map for 204
deglyco %>%
  filter(Ion == 204.087, 
         Exp == "- Glycosidase") %>%
  ggplot(aes(x = RT, 
             y = mz)) + 
  geom_point(aes(alpha = 4*Scaled.Intensity),
             size = 0.4, 
             shape = 15) +
  theme_linedraw() + 
  theme(strip.background = element_blank(),
        strip.text = element_text(colour = "black", face = "bold", size = 14),
        panel.grid = element_blank(),
        strip.text.x = element_text(hjust = -0.01),
        axis.title = element_text(size = 13),
        plot.margin = unit(c(1, 1, 1, 1), "cm"),
        ) +
  scale_alpha_identity() +
  scale_y_continuous(expand = expand_scale(mult = c(0, 0)),
                     limits = c(999,1501),
                     breaks = seq(1000,1500,by=100)) +  
  scale_x_continuous(expand = expand_scale(mult = c(0, 0)),
                     limits = c(5,45)) +
  ylab("Precursor m/z") +
  xlab("Retention Time") 

#ggsave("/Users/whitem/Documents/GitHub/OxoScan-MS/paper_analysis/FinalFigurePDFs/Figure2b.jpeg", height = 5, width = 7, dpi=600)


  

```

#### Oxonium map for all 3 ions - Figure S1b

``` {R ED_Figure1c_OxoniumMap, fid.width = 12.5, fig.height = 10, eval = T, echo = T, results = F}

## Format longer, remove zero values, scale and add labels by each ion 
deglyco.all <- deglyco.raw %>%
  filter(Exp == "- Glycosidase") %>% 
  mutate(mz = ((Window.Low+Window.High)/2),
          .keep = "unused",
          .before = 1) %>% ## Use middle of each m/z bin for precursor m/z plotting
  pivot_longer(cols = 3:11, 
               names_to = "Ion", 
               values_to = "Intensity") %>% ## Make 'long' table
  filter(Intensity != "0") %>% ## Remove 0 values
  group_by(Ion) %>% ## Scale intensity across control/deglycosidase experiments by each ion
  mutate(Scaled.Intensity = Intensity/max(Intensity)) %>%
  ungroup() 


## Plot faceted for each ion
deglyco.all %>%
  ggplot(aes(x = RT, 
             y = mz)) + 
  geom_point(aes(alpha = 4*Scaled.Intensity),
             size = 0.4, 
             shape = 15) +
  theme_linedraw() + 
  theme(strip.background = element_blank(),
        strip.text = element_text(colour = "black", face = "bold", size = 12, hjust = -0.01),        
        axis.title = element_text(size = 13),
        panel.grid = element_blank(),
        plot.margin = unit(c(1, 1, 1, 1), "cm"),
        ) +
  scale_alpha_identity() +
  scale_y_continuous(expand = expand_scale(mult = c(0, 0)),
                     limits = c(999,1501),
                     breaks = seq(1000,1500,by=100)) +  
  scale_x_continuous(expand = expand_scale(mult = c(0, 0)),
                     limits = c(5,45)) +
  ylab("Precursor m/z") +
  xlab("Retention Time") +
  facet_wrap(~paste("m/z", Ion), ncol = 3)


#write_csv(deglyco.all %>%
#            select(mz, RT, Ion, Intensity), "SourceData/FigureS1c.csv")

#ggsave("/Users/whitem/Documents/GitHub/OxoScan-MS/paper_analysis/FinalFigurePDFs/EDFigure1c.png", height = 8.5, width = 12.5, dpi=600)


```

#### Effect of treatment with glycosidase mix (16h) - Figure 2c

``` {R Figure2c_2x2, eval = T, echo = F}

## Faceted plot of plasma samples +/- glycosidase treatment, scaled by ion only
deglyco %>%
  filter(Ion != "274.092") %>% ## For the purpose of the plot, 274 shows same trend
  ggplot(aes(x = RT, 
             y = mz)) + 
  geom_point(aes(alpha = 1.5*Scaled.Intensity),
             size = 0.3, 
             shape = 15) +
  theme_linedraw() + 
  theme(strip.background = element_blank(),
        strip.text = element_text(colour = "black", face = "bold", size = 12),
        axis.title = element_text(size = 13),
        panel.grid = element_blank(),
        strip.text.x = element_text(hjust = -0.01),
        plot.margin = unit(c(0.1, 0.1, 0.1, 0.1), "cm")) +
  scale_alpha_identity() +
  scale_y_continuous(expand = expand_scale(mult = c(0, 0))) +  
  scale_x_continuous(expand = expand_scale(mult = c(0, 0)),
                     limits = c(5,45)) +
  ylab("Precursor m/z") +
  xlab("Retention Time") +
  facet_grid(Exp~paste("m/z = ", Ion, "\n", Ion.Label,sep="")) 

#ggsave("/Users/whitem/Documents/GitHub/OxoScan-MS/paper_analysis/FinalFigurePDFs/Figure2c.jpeg", height = 5, width = 7, dpi=600)
 
```

#### ~99% reduction in signal intensity upon deglycosidase treatment

``` {R Signal intensity across treatments, echo = F, eval = T}

### total signal in glyco/deglyco
deglyco %>%
  group_by(Ion, Exp) %>% ## Sum of oxonium intensities for each ion +/- glycosidase treatment
  summarise(sum = sum(Intensity), .groups = "drop_last") %>% ## Non-scaled intensity used
  pivot_wider(values_from = sum,
              names_from = Exp) %>%
  summarise(`% Signal Deglycosylated/Control` = 100 * `+ Glycosidase` /  `- Glycosidase`,
            .groups = "drop") ## %age signal in deglycosylated sample compared to untreated

```
### Comparison of repeat injections of a plasma tryptic digest

``` {R Figure2d_Correlation, fig.width = 5, fig.height = 5}

rep.df <- dir_ls(raw.data.dir,
                regexp = "TwoReplicates") %>%
  fread() %>%
  pivot_wider(values_from = "value",
              names_from = rep) %>%
  dplyr::rename('Replicate 1' = `2`,
         'Replicate 2' = `3`)

## Colour points by density - use function below (from: https://slowkow.com/notes/ggplot2-color-by-density/)

get_density <- function(x, y, ...) {
  dens <- MASS::kde2d(x, y, ...)
  ix <- findInterval(x, dens$x)
  iy <- findInterval(y, dens$y)
  ii <- cbind(ix, iy)
  return(dens$z[ii])
}


rep.df$density <- get_density(log10(rep.df$`Replicate 1`), log10(rep.df$`Replicate 2`), 
                              n = 5000,
                              h = c(0.1,0.1))


### Calculate spearman's rank correlation coefficient - no assumption of normality (non-parametric)

corr.coeff <- cor(rep.df$`Replicate 2`, rep.df$`Replicate 1`, method = "spearman")

col.scale <- c(min(rep.df$density), ((max(rep.df$density) + min(rep.df$density))/2), max(rep.df$density)) ## Define colour scale by density

col <- colorRamp2(breaks = col.scale, colors = c("dark blue", "light blue", "light yellow")) ## Make scale

rep.df %>%
  ggplot(aes(x = log10(`Replicate 1`),
             y = log10(`Replicate 2`),
             colour = density
             )) +
  geom_point(size = 2,
             alpha = 0.6,
             shape = 19) +
  theme_linedraw() + 
  scale_colour_gradient2(low = "dark blue",
                         mid = "light yellow",
                         high = "dark red", 
                         midpoint = ((max(rep.df$density) + min(rep.df$density))/2)) +
  theme(panel.grid = element_blank(),
        legend.position = "none",        
        axis.title = element_text(size = 13)) + 
  annotate("text",
           x = 2, y = 3.5, 
           parse = T,
           label = expression(rho~"= 0.995"),
           colour = "black", 
           size = 5,
           fontface = "bold") +
  xlab("log10(Replicate 1)") + 
  ylab("log10(Replicate 2)")

#write_csv(rep.df, "Figure2d.csv")
#ggsave("/Users/whitem/Documents/GitHub/OxoScan-MS/paper_analysis/FinalFigurePDFs/Figure2d.pdf", height = 5, width = 5)


```




### Quantitative performance benchmark - Plasma spike-in to E. coli background

``` {r Figure2e_SpikeBothPlots, fig.height = 5, fig.width = 10}

spike.raw <- dir_ls(raw.data.dir,
       regexp = "20220221_Spike_") %>%
  fread() %>%
  mutate(exp = sub(spectrum_name,
                   pattern = "20210723_glyco_plasma_ecoli_spike_2ug_(.).wiff.dia.extracted.txt.*",
                   replacement = "\\1")) %>%
  filter(exp %in% c(2:4,6)) %>%
  left_join(data.frame(
    exp = as.character(c(2:4,6)), ## Match up experiment labels
    Plasma = c(5,20,40,80),
    Ecoli = c(95,80,60,20))) %>% ## Plasma / E.coli proportions
  group_by(Plasma, peak_num) %>%
  mutate(Intensity = sum(value)) %>% ### Sum oxonium ion intensities for each feature (in each sample)
  ungroup() %>%
  select(Plasma, Ecoli, peak_num, merged_rt, merged_mz, persistence, aligned_rt,  aligned_mz, Intensity) %>%
  distinct()
  
### Read in E.coli only sample to use as exclusion list (i.e. quantify only plasma features for spike-in)
ecoli.ref <- dir_ls(raw.data.dir, 
                    regexp = "20220515_.*EColi") %>%
  fread() %>%
  filter(ion == 138.055) %>% ## single value per peak_num
  mutate(unique.id = paste(merged_rt, merged_mz, sep = "_")) %>% ## make RT/mz coordinate for each peak to search for in spiked experiment
  pull(unique.id) %>%
  unique()

### filter with 0.25min tolerance for E.coli peaks in spike-in data -- use to exclude E.coli features from quant analysis
ecoli.exclude <- data.frame()


for(i in seq_along(ecoli.ref)){

  ref_rt <- ecoli.ref[i] %>% ## pull out reference RT
    sub(pattern = "(.*)_\\d*",
        replacement = "\\1") %>%
    as.numeric()
  
  ref_mz <- ecoli.ref[i] %>% ## pull out reference mz
    sub(pattern = ".*_(\\d*)",
        replacement = "\\1") %>%
    as.numeric()
  
  ## search (with +/- 0.2 minute RT tolerance for feature in spike-in data)
  ecoli.exclude <- rbind(ecoli.exclude,
                         spike.raw %>%
                           filter(merged_rt > ref_rt - 0.3, ## with 0.3 min tolerance
                                  merged_rt < ref_rt + 0.3,
                                  merged_mz == ref_mz)) ## no m/z tolerance, should not change

}


### add in some filtering steps for the peaks
spike.filtered <- spike.raw %>% 
  filter(!peak_num %in% ecoli.exclude$peak_num) 


#### how many peaks - checkpoint
paste(unique(spike.filtered$peak_num) %>% unique() %>% length(),
      "filtered plasma peaks used for quant out of",
      unique(spike.raw$peak_num) %>% unique() %>% length(),
      "total called")


### calculate fold-changes to reference (dilution = 1)
spike.fc <- spike.filtered %>%
  select(peak_num, Plasma, Intensity) %>%
  pivot_wider(names_from = Plasma,
              values_from = Intensity) %>%
  mutate(`Dilution.0.5` = `80`/`40`, ## 80% plasma compared to ref (40% plasma) - theoretical dilution (from ref) = 0.5
         `Dilution.2` = `20`/`40`, ## 20% plasma compared to ref (40% plasma) - theoretical dilution (from ref) = 2
         `Dilution.8` = `5`/`40`) %>% ## 5% plasma compared to ref (40% plasma) - theoretical dilution (from ref) = 8
  select(peak_num, `40`, `Dilution.0.5`, `Dilution.2`, `Dilution.8`) %>%
  dplyr::rename(Ref.Intensity = "40") %>%
  pivot_longer(cols = 3:5,
               names_to = "Dilution.Factor",
               values_to = "Fold-change") %>%
  mutate(Dilution.Factor = sub(Dilution.Factor,
                        pattern = "Dilution\\.",
                        replacement = ""),
         Dilution.Factor = factor(Dilution.Factor, levels = c('0.5', '2', '8'))) %>% ## No pun intended
  filter(Ref.Intensity > 1) ## remove extreme values


spike.withboxplot <- plot_grid(
  spike.fc %>%
    ggplot(aes(x = log2(Ref.Intensity),
               y = log2(`Fold-change`),
               colour = Dilution.Factor)) + 
    geom_point(alpha = 0.75,
               size = 2.5,
               shape = 20) + 
    theme_linedraw() +
    theme(panel.grid = element_blank(),
          legend.position = "none",        
          axis.title = element_text(size = 13),
          plot.margin = unit(c(5.5, 0, 5.5, 0), "pt")) + 
    scale_colour_manual(values = pnw_palette("Sunset2", 3)) +
    scale_y_continuous(breaks = seq(-10,4,by=2),
                       limits = c(-6,4)) +
    scale_x_continuous(breaks = seq(0,16,by=4),
                       limits = c(2,15)) +
    geom_hline(yintercept = 1,
                colour = pnw_palette("Sunset2", 3)[1],
                linetype = "dashed") +  
    geom_hline(yintercept = -1,
                colour = pnw_palette("Sunset2", 3)[2],
                linetype = "dashed") +   
    geom_hline(yintercept = -3,
                colour = pnw_palette("Sunset2", 3)[3],
                linetype = "dashed") + 
    guides(colour=guide_legend(title="Dilution\nFactor")) +
    ylab("log2(Fold-Change)") + 
    xlab("log2(Reference Intensity)") 
  
    ,
  
  spike.fc %>%
    ggplot(aes(x=Dilution.Factor,
             y=log2(`Fold-change`),
             fill = Dilution.Factor)) + 
    geom_boxplot(alpha=0.85,
                 outlier.alpha = 0) + 
    geom_hline(yintercept = 1,
                colour = pnw_palette("Sunset2", 3)[1],
                linetype = "dashed") +  
    geom_hline(yintercept = -1,
                colour = pnw_palette("Sunset2", 3)[2],
                linetype = "dashed") +   
    geom_hline(yintercept = -3,
                colour = pnw_palette("Sunset2", 3)[3],
                linetype = "dashed") +
    theme_linedraw() + 
    scale_y_continuous(breaks = seq(-10,4,by=2),
                       limits = c(-6,4)) +
    ylab("") +
    xlab("log2(Reference Intensity)") +
    scale_fill_manual(values = pnw_palette("Sunset2", 3)) +
    guides(fill=guide_legend(title="Dilution\nFactor")) +
    theme(panel.grid = element_blank(),
          axis.text.x = element_text(colour = "white"),
          axis.text.y = element_blank(),        
          axis.ticks = element_blank(),
          axis.title.x = element_text(colour = "white"),
          axis.line = element_blank(),
          plot.margin = unit(c(5.5, 0, 5.5, 0), "pt"))
  ,
  ncol = 2,
  rel_widths = c(0.7,0.3)) 

spike.withboxplot

#write_csv(spike.fc, "SourceData/Figure2e.csv")
#ggsave("/Users/whitem/Documents/GitHub/OxoScan-MS/paper_analysis/FinalFigurePDFs/Figure2e.pdf", height = 5, width = 10)

```
